Embedding to Reference t-SNE Space Addresses Batch Effects in Single-Cell Classification

### 2019-08-13 paper/policar2019-tsne-embedding.tex
Shorten manuscript

Dimensionality reduction techniques, such as t-SNE, can construct informative
visualizations of high-dimensional data. When working with multiple data
sets, a straightforward application of these methods often fails; instead of
revealing underlying classes, the resulting visualizations expose data
set-specific clusters. To circumvent these batch effects, we propose an
embedding procedure that takes a t-SNE visualization constructed on a
reference data set and uses it as a scaffold for embedding new data. The new,
secondary data is embedded one data point at the time. This prevents any
interactions between instances in the secondary data and implicitly mitigates
batch effects. We demonstrate the utility of this approach by analyzing six
recently published single-cell gene expression data sets with up to tens of
thousands of cells and thousands of genes. The batch effects are particularly
strong as the used data comes from different institutions and was obtained
using different experimental protocols. The visualizations constructed by our
proposed approach are cleared of batch effects, and the cells from secondary
data sets correctly co-cluster with cells of the same type from the primary
data.

\keywords{Batch effects \and Embedding \and t-SNE \and Visualisation \and Single-Cell Transcriptomics \and Data Integration \and Domain Adaptation.}
\end{abstract}

@@

Massive, multi-variate biological data sets often suffer from these
source-specific biases.  The focus of this work is single-cell genomics, a
domain that was selected due high biomedical relevance and and abundance of
recently published data. Single-cell RNA sequencing (scRNA-seq) data sets are
the result of isolating RNA molecules from individual cells, which serve as an
estimate of the expression of cell's genes.  The studies can exceed thousands
of cells and tens of thousands of genes, and typically start with cell type
analysis. Here, it is expected that cells of the same type would cluster
together in two-dimensional data visualisation~\cite{seurat}. For instance,
Fig.~\ref{fig:batch_effect}.a shows t-SNE embedded data from mouse brain cells
originating from the visual cortex~\cite{hrvatin2018} and the
hypothalamus~\cite{chen2017}. The figure reveals distinct clusters but also
separates the data from the two brain regions. These two regions share the same
cell types and --- contrary to the depiction in Fig.~\ref{fig:batch_effect}.a
--- we would expect the data points from the two studies to overlap. Batch
effects similarly prohibit the utility of t-SNE in the exploration of
pancreatic cells in Fig.~\ref{fig:batch_effect}.b, which renders the data from
a pancreatic cell atlas~\cite{baron2016} and similarly-typed cells from
diabetic patients~\cite{xin2016}. Just like with data from brain cells,
pancreatic cells cluster primarily by data source, again resulting in a
visualization driven by batch effects.

\begin{figure}[htbp]
  \includegraphics[width=\textwidth]{figures/batch_effect.pdf}
  \caption{Batch effects are a driving factor of variation between the data
  sets.  Depicted is a t-SNE visualisation of two pairs of data sets. In each
  pair, the data sets share cell types, so it would be expected that the cells
  from the reference data (blue) would mix with the cells in a secondary data
  sets (orange).  Instead, t-SNE visualisation clusters data according to the
  data source.}
  \label{fig:batch_effect}
\end{figure}

@@

We propose a direct solution of rendering t-SNE visualizations that addresses
batch effects. Our approach treats one of the data sets as a {\em reference}
and aims to embed the cells from another, {\em secondary data set} to a common
low-dimensional space.  We construct a t\nobreakdash -SNE embedding using the
reference data set, and then use it as a scaffold for the embedding of the
secondary data.  The key idea underpinning our approach is that secondary data
points are embedded independently of one another.

@@

The proposed solution is implemented using a mapping of new data to an existing
t-SNE visualization. While the utility of such an algorithm was already hinted
at in recent publication~\cite{art_of_using_tsne}, here we provide its
practical and theoretically-grounded implementation. Considering the abundance
of recent publications on batch effect removal, we present surprising evidence
that a computationally more direct and principled embedding procedure solves
the batch effects problem when constructing interpretable visualizations from
different data sources.

@@

We describe an end-to-end pipeline that uses fixed t-SNE coordinates as a
scaffold for embedding new (secondary) data, enabling joint visualisation of
multiple data sources while mitigating batch effects. Our proposed approach
starts by using t\nobreakdash -SNE to embed a reference data set, with the aim
of constructing a two-dimensional visualisation to facilitate interpretation
and cluster classification. Then, the placement of each new sample is optimized
independently via the t\nobreakdash -SNE loss function. Independent treatment
of each data instance from a secondary data set disregards any interactions
present in that data set, and prevents the formation of clusters that would be
specific to the secondary data. Below, we start with a summary of t-SNE and its
extensions (Sec.~\ref{sec:tsne}, introducing the relevant notation, upon which
we base our secondary data embedding approach (Sec.~\ref{sec:transfer}).

@@

Local, non-linear dimensionality reduction by t-SNE is performed as follows.
Given a multi-dimensional data set $\mathbf{X} = \left \{ \mathbf{x}_1,
\mathbf{x}_2, \dots, \mathbf{x}_N \right \} \in \mathbb{R}^D$ where $N$ is the
number of data points in the reference data set, t-SNE aims to find a low
dimensional embedding $\mathbf{Y} = \left \{ \mathbf{y}_1, \mathbf{y}_2, \dots,
\mathbf{y}_N \right \} \\\in \mathbb{R}^d$ where $d \ll D$, such that if points
$\mathbf{x}_i$ and $\mathbf{x}_j$ are close in the multi-dimensional space,
their corresponding embeddings $\mathbf{y}_i$ and $\mathbf{y}_j$ are also
close. Since t-SNE is primarily used as a visualization tool, $d$ is typically
set to two. The similarity between two data points in t-SNE is defined as:

@@

The time complexity needed to evaluate the similarities in
Eq.~\ref{eq:cauchy_kernel} is $\mathcal{O}(N^2)$, making the application of the
algorithm impractical for large data sets.  Hence, we adopt a recent approach
for low-rank approximation of gradients based on polynomial interpolation which
reduces the time complexity of t-SNE to $\mathcal{O}(N)$. This approximation
enables the visualization of massive data sets, possibly containing millions of
data points~\cite{fi_tsne}.

The resulting embeddings substantially depend on the value of the perplexity
parameter. Perplexity can be interpreted as the number of neighbors for which
the distances in the embedding space are preserved. Small values of perplexity
result in tightly-packed clusters of points and effectively ignore the
long-range interactions between clusters. Larger values may result in a more
globally consistent visualisations --- preserving distances on a large scale
and organizing clusters in a more meaningful way --- but can lead to merging
small clusters and thus obscuring local aspects of the
data~\cite{art_of_using_tsne}.

@@

\noindent where $L$ is the number of mixture components. The bandwidths
$\sigma_{i,l}$ are selected in the same manner as in
Eq.~\ref{eq:gaussian_kernel}, but with a different value of perplexity for each
$l$. In our experiments, we used a mixture of two Gaussian kernels with
perplexity values of 50 and 500. We note that a similar formulation of
multi-scale kernels was proposed in~\cite{art_of_using_tsne}, and we found the
resulting embeddings are visually very similar to those obtained with the
approach described above (not shown for brevity).

@@

\noindent where $\mathbf{V} = \left \{ \mathbf{v}_1, \mathbf{v}_2, \dots,
\mathbf{v}_M \right \} \in \mathbb{R}^D$ where $M$ is the number of samples in
the new data set and $\mathbf{W} = \left \{ \mathbf{w}_1, \mathbf{w}_2, \dots,
\mathbf{w}_M \right \} \in \mathbb{R}^d$. Additionally, we omit the
symmetrization step in Eq.~\ref{eq:symmetrize}. This enables new points to be
inserted into the embedding independently of one another. The gradients of
$\mathbf{w}_j$ with respect to the loss (Eq.~\ref{eq:kl_divergence}) are:

@@

We use three pairs of reference and secondary single-cell data sets originating
from different organisms and tissues. The data in each pair were chosen so that
the majority of cell types from the secondary data set were included in the
reference set (Table~\ref{tab:data sets}). The cells in the data sets
originates from the following three tissues:

@@

\item[Human pancreas.] The data set from Baron \etal~\cite{baron2016} was
created as an atlas of pancreatic cell types. We used this set as a reference
for data from Xin \etal~\cite{xin2016}, who examined transcriptional
differences between healthy and type 2 diabetic patients.

\item[Mouse retina.] The data set from Macosko \etal~\cite{macosko2015} was
created as an atlas of mouse retinal cell types. We used this as a reference
for the data from Shekhar \etal~\cite{shekhar2016}, who built an atlas for
retinal bipolar cells.
\end{description}

@@

\noindent The mean $\log_2$ expression of the genes is computed from all the
cells where gene was expressed:

@@

Figs.~\ref{fig:transform_brain}, \ref{fig:transform_pancreas}, and
\ref{fig:transform_retina} show the embeddings of the reference data sets and
their corresponding embeddings of the secondary data sets. In all the figures,
the cells from the secondary data sets were positioned in the cluster of
same-typed reference cells, providing strong evidence of the success of the
proposed approach. There are some deviations to these observations; for
instance, in Fig.~\ref{fig:transform_brain} several oligodendrocyte precursor
cells (OPCs) were mapped to oligodendrocytes. This may be due to differences in
annotation criteria by different authors, or due to inherent similarities of
these types of cells. Examples of such erroneous placements can be found in
other figures as well, but they are not common and constitute less then 5\% of
the cells (less than  5\% in brain, \%1 in pancreas and \%2 in retina
secondary data sets).

@@

Notice that we could simulate the split between reference and secondary data
sets using one data set only and perform cross-validation, however this type of
experiment would not incorporate batch effects. We want to remind the reader
that handling batch effects were central to our endeavour and that the
disregard of this effect could lead to overly-optimistic results and data
visualizations strikingly different from ours. For example, compare the
visualisations from Fig.~\ref{fig:batch_effect}.a and
Fig.~\ref{fig:transform_brain}.b, or Figs.~\ref{fig:batch_effect}.b and
\ref{fig:transform_pancreas}.b.

@@

There are additional modifications that we use in the embedding of the
secondary data set that were recently proposed and enhance the original t-SNE
visualization.  One important extension is the use of multi-scale similarities
that, besides local ordering of the data points, includes global optimization
of cluster placement. For illustration, consider visualizations with standard
and multi-scale t-SNE in Fig.~\ref{fig:multiscale}. Notice, for instance, that
in multi-scale t-SNE (Fig.~\ref{fig:multiscale}.b) the clusters with neuronal
cells are clumped together, while their placement in standard t-SNE is
arbitrary (Fig.~\ref{fig:multiscale}.a).

@@

The proposed method assumes that all cell types from the secondary data set are
present in the reference. The proposed method would fail to reveal novel cell
types in the secondary data set, possibly positioning them arbitrarily close to
unrelated clusters. Procedures such as scmap were recently proposed to cope
with such cases and identify the cells whose type is new and not included in
the reference~\cite{scmap}.

Our procedure and the resulting cell landscape are therefore affectected by the
asymmetry of determining the reference and the secondary data sets. In
practice, however, newly produced \em secondary \em data would be embedded into
published and annotated reference landscapes. Scaling-up to a wider collection
of cell types is driven by increasingly large collections of data, e.g the
Human Cell Atlas initiative~\cite{hca}.

@@

The procedures described in this paper are provided as Python notebooks that
are, together with the data, available in an open
repository~\footnote{https://github.com/biolab/tsne-embedding}. All experiments
were run using openTSNE~\footnote{https://github.com/pavlin-policar/openTSNE},
our open and extensible t-SNE library for Python.

@@

Almost all recent publications of single-cell studies begin with a
two-dimensional visualization of the data that reveals the cellular diversity.
While most dimensionality reduction techniques can be used to render such a
visualization, different variants of t-SNE are most often used. Due to the
ability to explore biological mechanisms at the individual cell level,
single-cell studies are increasingly widespread, and their publications in the
past couple of years are abundant. One of the central tasks in single-cell
studies is the classification of new cells based on findings from previous
studies. Such transfer of knowledge is often difficult due to batch effects
present in data from different sources.  Addressing batch effects by adapting
and extending t-SNE, the prevailing method used to present single-cell data in
two-dimensional visualization motivated the research presented in this paper.

@@

This work was supported by the Slovenian Research Agency Program Grant P2-0209,
and by the BioPharm.SI project supported from European Regional Development
Fund and the Slovenian Ministry of Education, Science and Sport. We would also
like to thank Dmitry Kobak for discussions on t-SNE.

>>>>

Dimensionality reduction techniques, such as t-SNE, can construct informative
visualizations of high-dimensional data. When working with multiple data
sets, a straightforward application of these methods often fails; instead of
revealing underlying classes, the resulting visualizations expose data
set-specific clusters. To circumvent these batch effects, we propose an
embedding procedure that uses a t-SNE visualization constructed on a
reference data set as a scaffold for embedding new data points. The new,
secondary data is embedded one data point at the time. This prevents any
interactions between instances in the secondary data and implicitly mitigates
batch effects. We demonstrate the utility of this approach by analyzing six
recently published single-cell gene expression data sets with up to tens of
thousands of cells and thousands of genes. The batch effects are particularly
strong as the used data comes from different institutions and was obtained
using different experimental protocols. The visualizations constructed by our
proposed approach are cleared of batch effects, and the cells from secondary
data sets correctly co-cluster with cells of the same type from the primary
data.

\keywords{Batch effects \and Embedding \and t-SNE \and Visualization \and Single-Cell Transcriptomics \and Data Integration \and Domain Adaptation.}
\end{abstract}

@@

Massive, multi-variate biological data sets often suffer from these
source-specific biases.  The focus of this work is single-cell genomics, a
domain that was selected due high biomedical relevance and and abundance of
recently published data. Single-cell RNA sequencing (scRNA-seq) data sets are
the result of isolating RNA molecules from individual cells, which serve as an
estimate of the expression of cell's genes.  The studies can exceed thousands
of cells and tens of thousands of genes, and typically start with cell type
analysis. Here, it is expected that cells of the same type would cluster
together in two-dimensional data visualization~\cite{seurat}. For instance,
Fig.~\ref{fig:batch_effect}.a shows t-SNE embedded data from mouse brain cells
originating from the visual cortex~\cite{hrvatin2018} and the
hypothalamus~\cite{chen2017}. The figure reveals distinct clusters but also
separates the data from the two brain regions. These two regions share the same
cell types and --- contrary to the depiction in Fig.~\ref{fig:batch_effect}.a
--- we would expect the data points from the two studies to overlap. Batch
effects similarly prohibit the utility of t-SNE in the exploration of
pancreatic cells in Fig.~\ref{fig:batch_effect}.b, which renders the data from
a pancreatic cell atlas~\cite{baron2016} and similarly-typed cells from
diabetic patients~\cite{xin2016}. Just like with data from brain cells,
pancreatic cells cluster primarily by data source, again resulting in a
visualization driven by batch effects.

\begin{figure}[htbp]
  \includegraphics[width=\textwidth]{figures/batch_effect.pdf}
  \caption{Batch effects are a driving factor of variation between the data
  sets. We depict a t-SNE visualization of two pairs of data sets. In each
  pair, the data sets share cell types, so we would expect cells
  from the reference data (blue) to mix with the cells in a secondary data
  sets (orange). Instead, t-SNE clusters data according to the
  data source.}
  \label{fig:batch_effect}
\end{figure}

@@

We propose a direct solution of rendering t-SNE visualizations to address
batch effects. Our approach treats one of the data sets as a {\em reference}
and embeds the cells from another, {\em secondary data set} to a joint
low-dimensional space.  We construct a t\nobreakdash -SNE embedding using the
reference data set, which is then used as a scaffold to embed the
secondary data. The key idea underpinning our approach is that secondary data
points are embedded independently of one another.

@@

The proposed solution implements a mapping of new data into an existing
t-SNE visualization. While the utility of such an algorithm was already hinted
at in recent publication~\cite{art_of_using_tsne}, here we provide its
practical and theoretically-grounded implementation. Considering the abundance
of recent publications on batch effect removal, we present surprising evidence
that a computationally more direct and principled embedding procedure solves
the batch effects problem when constructing interpretable visualizations from
different data sources.

@@

We describe an end-to-end pipeline that uses fixed t-SNE coordinates as a
scaffold for embedding new (secondary) data, enabling joint visualization of
multiple data sources while mitigating batch effects. Our proposed approach
starts by using t\nobreakdash -SNE to embed a reference data set, with the aim
of constructing a two-dimensional visualization to facilitate interpretation
and cluster classification. Then, the placement of each new sample is optimized
independently via the t\nobreakdash -SNE loss function. Independent treatment
of each data instance from a secondary data set disregards any interactions
present in that data set, and prevents the formation of clusters that would be
specific to the secondary data. Below, we start with a summary of t-SNE and its
extensions (Sec.~\ref{sec:tsne}, introducing the relevant notation, upon which
we base our secondary data embedding approach (Sec.~\ref{sec:transfer}).

@@

Local, non-linear dimensionality reduction by t-SNE is performed as follows.
Given a multi-dimensional data set $\mathbf{X} = \left \{ \mathbf{x}_1,
\mathbf{x}_2, \dots, \mathbf{x}_N \right \} \in \mathbb{R}^D$ where $N$ is the
number of data points in the reference data set, t-SNE aims to find a low
dimensional embedding $\mathbf{Y} = \left \{ \mathbf{y}_1, \mathbf{y}_2, \dots,
\mathbf{y}_N \right \} \in \mathbb{R}^d$ where $d \ll D$, such that if points
$\mathbf{x}_i$ and $\mathbf{x}_j$ are close in the multi-dimensional space,
their corresponding embeddings $\mathbf{y}_i$ and $\mathbf{y}_j$ are also
close. Since t-SNE is primarily used as a visualization tool, $d$ is typically
set to two. The similarity between two data points in t-SNE is defined as:

@@

The time complexity needed to evaluate the similarities in
Eq.~\ref{eq:cauchy_kernel} is $\mathcal{O}(N^2)$, making its application
impractical for large data sets. We adopt a recent approach
for low-rank approximation of gradients based on polynomial interpolation which
reduces its time complexity to $\mathcal{O}(N)$. This approximation
enables the visualization of massive data sets, possibly containing millions of
data points~\cite{fi_tsne}.

The resulting embeddings substantially depend on the value of the perplexity
parameter. Perplexity can be interpreted as the number of neighbors for which
the distances in the embedding space are preserved. Small values of perplexity
result in tightly-packed clusters of points and effectively ignore the
long-range interactions between clusters. Larger values may result in a more
globally consistent visualizations --- preserving distances on a large scale
and organizing clusters in a more meaningful way --- but can lead to merging
small clusters and thus obscuring local aspects of the
data~\cite{art_of_using_tsne}.

@@

\noindent where $L$ is the number of mixture components. The bandwidths
$\sigma_{i,l}$ are selected in the same manner as in
Eq.~\ref{eq:gaussian_kernel}, but with a different value of perplexity for each
$l$. In our experiments, we used a mixture of two Gaussian kernels with
perplexity values of 50 and 500. A similar formulation of
multi-scale kernels was proposed in~\cite{art_of_using_tsne}, and we found the
resulting embeddings are visually very similar to those obtained with the
approach described above (not shown for brevity).

@@

\noindent where $\mathbf{V} = \left \{ \mathbf{v}_1, \mathbf{v}_2, \dots,
\mathbf{v}_M \right \} \in \mathbb{R}^D$ where $M$ is the number of samples in
the secondary data set and $\mathbf{W} = \left \{ \mathbf{w}_1, \mathbf{w}_2, \dots,
\mathbf{w}_M \right \} \in \mathbb{R}^d$. Additionally, we omit the
symmetrization step in Eq.~\ref{eq:symmetrize}. This enables new points to be
inserted into the embedding independently of one another. The gradients of
$\mathbf{w}_j$ with respect to the loss (Eq.~\ref{eq:kl_divergence}) are:

@@

We use three pairs of reference and secondary single-cell data sets originating
from different organisms and tissues. The data in each pair were chosen so that
the majority of cell types from the secondary data set were included in the
reference set (Table~\ref{tab:data sets}). The cells in the data sets
originate from the following three tissues:

@@

\item[Human pancreas.] Baron \etal~\cite{baron2016} created
an atlas of pancreatic cell types. We used this set as a reference
for data from Xin \etal~\cite{xin2016}, who examined transcriptional
differences between healthy and type 2 diabetic patients.

\item[Mouse retina.] Macosko \etal~\cite{macosko2015} created
an atlas of mouse retinal cell types. We used this as a reference
for the data from Shekhar \etal~\cite{shekhar2016}, who built an atlas for
retinal bipolar cells.
\end{description}

@@

\noindent The mean $\log_2$ expression of gene $g$ considers only cells
expressing $g$:

@@

Figs.~\ref{fig:transform_brain}, \ref{fig:transform_pancreas}, and
\ref{fig:transform_retina} show the embeddings of the reference data sets and
their corresponding embeddings of the secondary data sets. In all the figures,
the cells from the secondary data sets were positioned in the cluster of
same-typed reference cells, providing strong evidence of the success of our approach. There are some deviations to these observations; for
instance, in Fig.~\ref{fig:transform_brain} several oligodendrocyte precursor
cells (OPCs) were mapped to oligodendrocytes. This may be due to differences in
annotation criteria by different authors, or due to inherent similarities of
these types of cells. Examples of such erroneous placements can be found in
other figures as well, but are uncommon and constitute less then 5\% of
the cells (less than 5\% in brain, 1\% in pancreas and 2\% in retina
secondary data).

@@

Notice that we could simulate the split between reference and secondary data
sets using one data set only and perform cross-validation, however this type of
experiment would not incorporate batch effects. We want to remind the reader
that handling batch effects were central to our endeavour and that the
disregard of this effect could lead to overly-optimistic results and data
visualizations strikingly different from ours. For example, compare the
visualizations from Fig.~\ref{fig:batch_effect}.a and
Fig.~\ref{fig:transform_brain}.b, or Figs.~\ref{fig:batch_effect}.b and
\ref{fig:transform_pancreas}.b.

@@

We use a number of additional, recently proposed modifications to enhance the t-SNE visualization of the reference data set.
One important extension is the use of multi-scale similarities, which, in addition to finding local structure, also ensure consistent organization of the resulting clusters.
For illustration, consider visualizations with standard
and multi-scale t-SNE in Fig.~\ref{fig:multiscale}. Notice, for instance, that
in multi-scale t-SNE (Fig.~\ref{fig:multiscale}.b) the clusters with neuronal
cells are clumped together, while their placement in standard t-SNE is
arbitrary (Fig.~\ref{fig:multiscale}.a).

@@

Our approach assumes that all cell types from the secondary data set are
present in the reference. The method would fail to reveal novel cell
types in the secondary data set, possibly positioning them arbitrarily close to
unrelated clusters. Procedures such as scmap were recently proposed to cope
with such cases and identify the cells whose type is new and not included in
the reference~\cite{scmap}.

Our procedure is, therefore, asymmetrical in the choice of reference and secondary data set.
In practice, however, newly produced {\em secondary} data would be embedded into previously-prepared reference landscapes. Large collections of data {\em e.g.} the
Human Cell Atlas initiative~\cite{hca} make it possible to scale up our approach to wider sets of cell types.

@@

The procedures described in this paper are provided as Python notebooks that
are, together with the data, available in an open
repository\footnote{https://github.com/biolab/tsne-embedding}. All experiments
were run using {\tt openTSNE},
our open and extensible t-SNE library for Python~\cite{opentsne}.

@@

Almost all recent publications of single-cell studies begin with a
two-dimensional visualization of the data that reveals cellular diversity.
While many dimensionality reduction techniques are available, different 
variants of t-SNE are most often used to produce such
visualizations. Single-cell studies enable the exploration of biological mechanisms at a
cellular level, and their publications in the past couple of years are abundant.
One of the central tasks in single-cell
studies is the classification of new cells based on findings from previous
studies. Such transfer of knowledge is often difficult due to batch effects
present in data from different sources. Addressing batch effects by adapting
and extending t-SNE, the prevailing method used to present single-cell data in
two-dimensional visualization, motivated the research presented in this paper.

@@

This work was supported by the Slovenian Research Agency Program Grant P2-0209,
and by the BioPharm.SI project supported from European Regional Development
Fund and the Slovenian Ministry of Education, Science and Sport. We would also
like to thank Dmitry Kobak for many helpful discussions on t-SNE.
